Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2020-02-10 ms/second-review-responses.tex

\begin{quote}
I am fairly pleased overall with the revised version of this article. I commend the authors for the substantial amount of work conducted. This contribution now convincingly demonstrates how important it is to properly account for population structure when individuals are distributed along a spatial continuum.
\end{quote}

\begin{point}{}
A couple of minor points about the forward model are still puzzling me though. First of all, in one dimension, the variance of $L(o)$, the random variable corresponding to the offspring's spatial position, given $l(p)$, the parent's position, is equal to $E((L(o)-l(p))^2):=\sigma^2$. Hence, $\sigma$ is here the square root of the expected squared Euclidean distance between parents and offspring. In general, assuming isotropic migrations, $\sigma$ is the square root of the expected squared Euclidean distance taken along any dimension. It seems to me that this definition is distinct from the mean parent-offspring distance, as stated by the authors.
\end{point}

\reply{
Good point here. We have corrected this to "$\sigma$ is the mean parent--offspring distance along each of the x and y axes" and now direct readers to \citep{Rousset1997} for discussion of how these parameters are defined in one- and two-dimensional habitats. The mean parent-offspring distance in two dimensions is $\sigma\sqrt{2}$, but we follow previous work in defining neighborhood size as $4 \pi \rho \sigma^2$ (i.e. without the factor of $\sqrt{2}$; see \citep{Rousset1997} for details). \llname{axialdist}
}

\begin{point}{}
More importantly, I am still a bit confused about the reproduction scheme. My understanding is that, once mating takes place, the two parents involved (noted as P(1) and P(2)) each produce a Poisson number of offspring. The location of each offspring is determined by a Gaussian distribution of covariance matrix $\sigma^2 I$, centered on the location of the parent that ``produced' this offspring. Let $O(1,i)$ and $O(2,j)$ be the i-th and j-th offspring produced by $P(1)$ and $P(2)$ respectively. Now, if I am not mistaken, half of the genetic material of O(2,j) comes from P(1) (the other half coming from P(2)) while O(1,i) receives one chromosome from P(2) (and another one from its ``true' parent, i.e., P(1)). Hence, when it comes to the parent-offspring distance as measured from their genetic material, half of the chromosomes have a ``small' dispersal distance, equal to that of the parent-offspring distance, while the other half have a ``large' distance roughly equal to the parent-offspring distance plus the distance between mates.

It is important that the authors clarify this last point because most results presented in this study rely on Wright's neighborhood size, which derives from a model where dispersal of individuals and chromosomes follow well-defined spatial dynamics. A sensible comparison of neighborhood sizes requires that dispersal (along with effective population density) is quantified in a coherent way across different models.

Additionally, the neighborhood size as defined by the authors (see line 211) is a function of the census density rather than the effective density of individuals. I am therefore not fully convinced that the range of values for this parameter is comparable to that given in Table 1.
\end{point}

\reply{
Another important point set of points on relating neighborhood sizes from different models here. In our simulation the average x axis displacement of a locus along any link in the pedigree is $\sigma$ for the "maternal" material (i.e. deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with variance $\sigma^2$. Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$. So from the perspective of genes sampled at the present time our neighborhood size calculations are lower than those in models like those described in \cite{Rousset1997} by a factor of (3/2) because of differences in the breeding model.

This inflation in neighborhood size is then partially offset by our use of census rather than "effective" population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly (2/3) the census size. We chose this method for density and sigma calculations because we sought to parameterize our model with values that could in theory be observed in real populations. We also find the concept of "effective population size" in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.

Thus relative to models like \cite{Rousset1997} our definition of \sigma is probably too low and \rho probably too high; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent. We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of \rho in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick. Indeed for many of the empirical studies the range of NS values given is larger than the differences we expect from the parameter definitions described above. We also think that our study of lattice simulations in appendix 1 demonstrates nicely how qualitatively similar simulations can lead to quite different patterns of genetic diversity at a given neighborhood size as a result of differences in the underlying breeding model.

\cjb{CJ: let's meet and talk again and talk about what edits to make here. I'm not sure what the best way to go is.}

@@

}

\begin{point}{}
Furthermore, the neighborhood size estimates obtained for Bebicium vittatum derive from the analysis of a one-dimensional habitat. The product $\sigma^2 \rho$ is thus expressed in this particular case as a number of individuals *per unit of space* (in two dimensions, this product is expressed as a number of individuals). Hence, the unit of space matters a lot here, prohibiting the direct comparison with other figures in the same table or with the simulations conducted in this study.
\end{point}

\reply{
Good point, we have removed Bebicium vittatum from table 1.  
}

### 2020-02-10 ms/spaceness.tex
expanded rev 3 responses

We simulated populations using the program SLiM v3.1 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance along each of the x and y axes.

@@

\subsection{Where are natural populations on this spectrum?}
For how much of the tree of life do spatial patterns circumscribe genomic variation?
In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms
to get an idea of how strongly local geographic dispersal affects patterns of variation.
This is an imperfect measure:
some aspects of genetic variation are most strongly determined by neighborhood size \citep{Wright1946},
others (e.g., number of segregating sites)
are more strongly determined by global $N_e$ or by the ratio of the two, and definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species \llabel{rhocaveat}. 
% These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$. 
\llabel{popsizecaveat}
In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density). However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants with neighborhood sizes less than roughly 100. Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation. 
Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A.~aegypti}). Last, many species likely have neighborhood sizes much larger than we simulated, including the recent ancestors of modern humans in northeastern Europe \citep{Ringbauer2017}. For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape. While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.

>>>>

We simulated populations using the program SLiM v3.1 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Rousset1997}, $\sigma$ is equal to the mean parent-offspring distance along each of the x and y axes.

@@

\subsection{Where are natural populations on this spectrum?}
For how much of the tree of life do spatial patterns circumscribe genomic variation?
In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms
to get an idea of how strongly local geographic dispersal affects patterns of variation.
This is an imperfect measure:
some aspects of genetic variation are most strongly determined by neighborhood size \citep{Wright1946},
others (e.g., number of segregating sites)
by global $N_e$, or the ratio of the two. 
\llabel{popsizecaveat}
In addition, definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species \llabel{rhocaveat}, and these empirical examples may be biased towards small-neighborhood species because few studies have quantified neighborhood size in species with very high dispersal or population density.

However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants with neighborhood sizes less than roughly 100. Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation. 
Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A.~aegypti}). Last, many species likely have neighborhood sizes much larger than we simulated, including the recent ancestors of modern humans in northeastern Europe \citep{Ringbauer2017}. For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape. While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.
